Asymmetric relationships and their effects on coexistence: Supporting information

### 2023-07-28 paper/ExplanationOfATAEffectsWorkingDoc.tex
wrote an intuition para in the Discussion and put the SI material in. Need to edit. Need to shorten and simplify the whole doc as much as possible.

\subsection{SI section X1 corresponding to the thoughts above, call it Temporal niche differentiation is necessary for ATAs}\label{SIsect:temporal_niche_diff}

@@

\subsection{SI section X2 corresponding to the thoughts above, call it Intuition and 
examples for ATAs}\label{SIsect:intuition_examples}

>>>>

\subsection{SI section X1 corresponding to the thoughts above, call it Temporal niche differentiation is necessary for ATA effects}\label{SIsect:temporal_niche_diff}

@@

\subsection{SI section X2 corresponding to the thoughts above, call it Intuition and 
examples for ATA effects}\label{SIsect:intuition_examples}

### 2023-07-28 paper/Paper.Rnw
wrote an intuition para in the Discussion and put the SI material in. Need to edit. Need to shorten and simplify the whole doc as much as possible.

%Para continuing to try to build intuition about how ATAs work
We return to the task, begun in the Introduction, of building intuition for how
ATA effects work. 
Just as for storage effects [\cite{Chesson_1994}, Box 2 of \cite{StumpVasseur2023}], 
temporal niche differentiation is necessary for ATA
effects (SI section X1). 
%DAN: fix reference
If temporal niche differentiation occurs
but is partial, as will be typical, $E_i$ and $C_{i \bs i}$ for a 
species $i$ will be positively but imperfectly associated, 
and the association may be asymmetric. 
Storage events occur for species $i$ when the positive effects of $E_i$ overwhelm the negative 
effects of $C_{i \bs i}$, producing large values of $r_i(E_i,C_{i \bs i})$
which are not completely offset by declines at other times. 
For instance, for the lottery model,
$r_i(E_i,C_{i \bs i})=\log(1-\delta+E_i/C_{i \bs i})$ can get large whenever
$C_{i \bs i}$ is a small fraction of $E_i$, but can never get smaller
than $\log(1-\delta)$. 
The potential for storage events can differ when $E_i$ and $C_{i \bs i}$ are both
above their median values compared to when they are both below median. For instance,
the ratio $E_i/C_{i \bs i}$, and therefore the growth rate 
$\log(1-\delta+E_i/C_{i \bs i})$, can both get very big, corresponding to
rapid growth, if $C_{i \bs i}$ is close to $0$, even if $E_i$ is fairly small.
In contrast, if $E_i$ and $C_{i \bs i}$ are both in the right portions of their distributions, 
so that $C_{i \bs i}$ is bounded away from $0$, there may be much sharper limits on the 
ratio $E_i/C_{i \bs i}$ and the corresponding growth rate.
ATA effects quantify the extent to which asymmetries of association between $E$ and 
$C$ are aligned with varying potential for storage across the $E,C$ distribution. 
ATA effects help quantify different degrees of temporal niche differentiation 
occurring under different sets of environmental conditions, possibly via
different mechanisms.
Details and examples of this intuition 
are in SI section X2, 
%DAN: Fix reference
where comparisons are also made between our beta and log-normal fecundities 
lottery models.

>>>>

%Para continuing to try to build intuition about how ATAs work
We return to the task, begun in the Introduction, of building intuition for how
ATA effects work. 
Just as for storage effects [\cite{Chesson_1994}, Box 2 of \cite{StumpVasseur2023}], 
temporal niche differentiation is necessary for ATA
effects (SI section \ref{SI-SIsect:temporal_niche_diff}). 
If temporal niche differentiation occurs
but is partial, as will be typical, $E_i$ and $C_{i \bs i}$ for a 
species $i$ will be positively but imperfectly associated, 
and the association may be asymmetric. 
Storage events occur for species $i$ when the positive effects of $E_i$ overwhelm the negative 
effects of $C_{i \bs i}$, producing large values of $r_i(E_i,C_{i \bs i})$
which are not completely offset by declines at other times. 
For instance, for the lottery model,
$r_i(E_i,C_{i \bs i})=\log(1-\delta+E_i/C_{i \bs i})$ can get large whenever
$C_{i \bs i}$ is a small fraction of $E_i$, but can never get smaller
than $\log(1-\delta)$. 
The potential for storage events can differ when $E_i$ and $C_{i \bs i}$ are both
above their median values compared to when they are both below median. For instance,
the ratio $E_i/C_{i \bs i}$, and therefore the growth rate 
$\log(1-\delta+E_i/C_{i \bs i})$, can both get very big, corresponding to
rapid growth, if $C_{i \bs i}$ is close to $0$, even if $E_i$ is fairly small.
In contrast, if $E_i$ and $C_{i \bs i}$ are both in the right portions of their distributions, 
so that $C_{i \bs i}$ is bounded away from $0$, there may be much sharper limits on the 
ratio $E_i/C_{i \bs i}$ and the corresponding growth rate.
ATA effects quantify the extent to which asymmetries of association between $E$ and 
$C$ are aligned with varying potential for storage across the $E,C$ distribution. 
ATA effects help quantify different degrees of temporal niche differentiation 
occurring under different sets of environmental conditions, possibly via
different mechanisms.
Details and examples of this intuition 
are in SI section \ref{SI-SIsect:intuition_examples} and SI 
Figs \ref{SI-fig:beta_fecundidities_explanatory}-\ref{SI-fig:lognormal_fecundidities_explanatory}, 
where comparisons are also made between our two lottery models.

### 2023-07-28 paper/SupMat.Rnw
wrote an intuition para in the Discussion and put the SI material in. Need to edit. Need to shorten and simplify the whole doc as much as possible.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLBqij.pdf}
\caption[Analogous to Fig. \ref{MT-fig:fig4} but making a different choice for $q_{ij}$]
{Analogous to Fig. \ref{MT-fig:fig4} of the main text, but using 
$q_{ij}=\eta_i/(\eta_j(1-\delta)+\eta_i \delta)$. See the caption of
that figure for details. Standard errors of plotted quantities 
were never greater than \Sexpr{round(se4qij,5)}, so are not displayed.}\label{fig:fig4_alt}
\end{figure}
%***DAN: Jasmin, the value of se4 quoted in the main text was the same, to 5 digits, as the 
%value of se4qij quoted here. This made me suspicious so I checked the values to 9 digits
%and they were the same to 9 digits. This seems like it must be an error in the code. Please
%check. The standard error expressions in SI S5.2 DO depend on qij, so this seems like it
%should not happen. The standard errors listed in Fig. 3 and Fig. S1 are different, so that's
%good.
%JASMIN: I think its okay that they are the same.

@@

\begin{figure}
\includegraphics[width=.66\textwidth]{../results_figs/VKQfig.pdf}
\caption[Maximum growth rate, cellular silicate content, and half sturation constant dependences on temperature.]
{Species maximum population growth rates, $V_i$ (a), half saturation constants for the dependence 
of growth rates on nutrient concentration, $K_i$ (b), and cellular silicate content, $Q_i$ (c) dependend on 
temperature in ways explored by \cite{Descamps_2005} and \cite{Ellner_2016, Ellner_2019} and 
pictured here. Species 1 is \emph{F. crotonensis}, shown in blue, and species 2 is 
\emph{C. pseudostelligera}, shown in red. Open circles and dashed lines indicate estimated 
values and interpolations from \cite{Descamps_2005} and closed circles and solid lines indicate 
values and functions used by \cite{Ellner_2016, Ellner_2019} and in our analyses. The two sets of
estimates and functions coincide on (c), except for the right-most point for species 2. For that species,
data were not available for $Q$ at $24^\circ$C from the original study of \cite{Descamps_2005}; \cite{Ellner_2016, Ellner_2019}
assumed constancy of $Q_2$ for temperature between $18^\circ$C and $24^\circ$C. Data were also not avaialable at
$24^\circ$C for $K_2$, and \cite{Ellner_2016, Ellner_2019} assumed constancy of $K_2$ for teperatures from 
$12^\circ$C through $24^\circ$C. \label{fig:VKQ}}
\end{figure}

>>>>

\section{   Temporal niche differentiation is necessary for ATA effects}\label{SIsect:temporal_niche_diff}

Just as for storage effects [see \cite{Chesson_1994} and Box 2 of \cite{StumpVasseur2023}], 
temporal niche differentiation is necessary in order for ATA
effects to materialize; we here explain why, in detail. ``Lack of temporal niche differentiation'' can be interpreted
to mean that species respond in identical ways to environmental fluctuations. 
In a two-species model such as those we consider, this should mean a monotonically
increasing relationship between $E_i$ and $C_{i \bs i}$. The former is the environment
as perceived by species $i$. The latter is competition that a scarce invader $i$ experiences
in a monoculture of $j$, and that competition comes only from $j$ (because $i$ itself
is scarce), so should correspond to the environmental response of $j$. If 
$E_i$ and $C_{i \bs i}$ are monotonically related, then $(E_i,C_{i \bs i})$
and $(E_i^\nat,C_{i \bs i}^\nat)$ are distributed in the same way, so 
$\E[r_i(E_i,c_{i \bs i})]-\E[r_i(E_i^\nat,c_{i \bs i}^\nat)]=0$.
Therefore, 
\begin{align}
\Delta_i^{[EC]} &= (\olr_{i \bs i}-\olr_{j \bs i})-(\olr_{i \bs i}^\nat - \olr_{j \bs i}^\nat) \\
&= \olr_{i \bs i}-(\olr_{i \bs i}^\nat - \olr_{j \bs i}^\nat) \\
&= \E[r_i(E_i,C_{i \bs i})]-\E[r_i(E_i^\nat,C_{i \bs i}^\nat)]+\E[r_j(E_j^\nat,C_{j \bs i}^\nat)] \label{eq:expri}\\
&= \E[r_j(E_j^\nat,C_{j \bs i}^\nat)].
\end{align}
But we already know $\E[r_j(E_j^\nat,C_{j \bs i}^\nat)]=0$
because there is no niche differentiation between $j$ and itself, so 
$E_j$ and $C_{j \bs i}$ should be monotonically related. So 
$\E[r_j(E_j^\nat,C_{j \bs i}^\nat)] = \E[r_j(E_j,C_{j \bs i})]=0$, giving
$\Delta_i^{[EC]}=0$. So temporal
niche differentiation, interpreted as Spearman correlation between 
$E$ and $C$ less than $1$, is necessary for ATA effects on coexistence to 
materialize.

\section{   Intuition and examples for ATA effects}\label{SIsect:intuition_examples}

If temporal niche differentiation occurs but is partial, 
then $E_i$ and $C_{i \bs i}$ will be positively but imperfectly correlated with
each other, corresponding to similar but non-identical responses of species to
environmental fluctuation (SI section \ref{SIsect:temporal_niche_diff}). 
But how do ATA effects then happen? We here extend the intuition
introduced in the Discussion and provide examples.

First, we illustrate using the beta fecundities lottery model that
the potential for storage to occur when 
$E$ and $C$ are both higher than average, versus the potential  
when $E$ and $C$ are
both lower than average, are not necessarily equivalent.
Storage can occur if population
increases during good years outweigh decreases during bad years. For instance,
if $r_i(E_i,C_{i \bs i})$ is very large when the balance of $E_i$ and 
$C_{i \bs i}$ is favorable, but only moderately small when this balance is
unfavorable, the expected value $\E[r_i(E_i,C_{i \bs i})]$ in (\ref{eq:expri})
will be positive. But the potential for such storage of population growth
may differ between when $E_i$ and $C_{i \bs i}$ are both high compared
to when they are both low. 
For instance, for the beta fecundities lottery model, $E_i/C_{i \bs i}$, a quantity
which features prominently in $r_i(E_i,C_{i \bs i})=\log(1-\delta+E_i/C_{i \bs i})$,
takes values bounded above by $\frac{2\eta_i \delta}{\eta_j}$ 
and bounded below by $\frac{\eta_i \delta}{2 \eta_j}$ when $E_i$ and $C_{i \bs i}$
are both in the upper halves of their assumed beta distributions (recall that 
these quantities were assumed to be distributed, respectively, as $\eta_1 \beta$
and $\eta_2 \beta/\delta$ for $\beta$ a beta-distributed random variable with parameters 
$0.5$ and $0.5$). These bounds are separated only by a factor of four. 
On the other hand, when $E_i$ and $C_{i \bs i}$ are both in the lower halves 
of their distributions, $E_i/C_{i \bs i}$ can take any positive value because
then either $E_i$ or $C_{i \bs i}$ can take values arbitrarily close to $0$. 
Because values of $E_i/C_{i \bs i}$ have the potential to be much larger when 
$E_i$ and $C_{i \bs i}$ are both in the lower halves of their distributions
than when these quantities are both in their upper halves, and because
the growth rate, $\log(1-\delta+E_i/C_{i \bs i})$, buffers low values of 
$E_i/C_{i \bs i}$ (the infemum of the growth rate is $\log(1-\delta)$, whereas the supremum is
infinite), the storage effects contributed by the left tails of 
$E_i$ and $C_{i \bs i}$ have the potential to be much greater than those contributed
by the right tails. Essentially, annual growth-when-rare rates for species $i$
that occur when $E_i$ and $C_{i \bs i}$ are both in the upper halves of their
distributions can never exceed $\log(1-\delta+\frac{2\eta_i \delta}{\eta_j})$,
whereas rates ocurring when $E_i$ and $C_{i \bs i}$ are both in the lower
halves of their distributions can be arbitrarily large. 
Apparently, the years with the most potential to 
contribute to storage
effects for the beta-fecundities lottery model are those for which the environment
is actually poorer than average, because, for those years, competition can be very low,
so that the ratio $E_i/C_{i \bs i}$ is large. Years for which the
environment is exceptionally good actually cannot contribute much to storage effects for 
this model, because for those years, $E_i/C_{i \bs i}$ cannot be very large.
It is the \emph{imbalance} between $E$ and $C$ that allows for storage effects,
not solely the quality of environment. 
Counterintuitively, that imbalance can be greater for this model when the environment is 
sub-optimal.

Fig. \ref{fig:beta_fecundidities_explanatory} provides quantitative backing to
the ideas of the prior paragraph, so we orient the reader 
to the figure before using it below to further explain
ATAs. The growth function $\ln(1-\delta+E/C)$ that pertains to the lottery model
is plotted with colors and contours on Fig. \ref{fig:beta_fecundidities_explanatory}A-C.
Samples were taken from the distribution $(E_i,C_{i \bs i})$ for the beta 
fecundities lottery model for left- and right-tailed 
associations between $E_i$ and $C_{i \bs i}$, and are plotted as
points on Fig. \ref{fig:beta_fecundidities_explanatory}A and C, respectively. 
Samples were also taken from the distribution $(E_i^\nat,C_{i \bs i}^\nat)$
(which is the same regardless of whether the original $(E_i,C_{i \bs i})$
distribution showed left- or right-tail association) and are plotted on 
Fig. \ref{fig:beta_fecundidities_explanatory}B. 
The dashed lines labelled $L_2$ on A-C separate the upper-right $50\%$
of each distribution from the lower-left $50\%$, and those portions are labelled. 
The quantity $\E \left[ \ln \left( 1-\delta+\frac{E_i}{C_{i \bs i}} \right) \right]$ can be
well approximated, for left- and right-tail association cases, by averaging values of the
function depicted with color and contours over the points on panels A and C, 
respectively. The quantity
$\E \left[ \ln \left( 1-\delta+\frac{E_i^{||}}{C_{i \bs i}^{||}} \right) \right]$
can be well approximated by averaging function 
values over the points on B. So ATA effects for the model with left-tail association
between $E_i$ and $C_{i \bs i}$ (respectively, right-tail association) are given by 
computing the average function value over points on panel A (respectively, C)
and then subtracting the average function value over points on panel B.
Contributions to the storage effect of points in the left portions of 
the $E,C$ distribution (the region labeled ``left'' on panels A and C)
can be accurately estimated by averaging function values across
points in that region; likewise storage contributions of points in the 
``right'' region can be estimated by averaging function values over 
points in that region.

The details of the averaging processes described above are pictured in 
Fig. \ref{fig:beta_fecundidities_explanatory}, panels D-F. 
Signed distances of points on Fig. \ref{fig:beta_fecundidities_explanatory} 
A-C from the lines $L_1$ correspond to 
the quantities $\log_{10}(E_1/C_{1 \bs 1})$ on A, C and to 
$\log_{10}(E_1^\nat/C_{1 \bs 1}^\nat)$ on B. The distributions of these 
quantities across points in the ``left'' region are pictured 
in the lower horizontal-axis histograms on panels D-F, respectively;
and distributions for points in the ``right'' region are 
pictured in the upper horizontal-axis histograms. 
``Right'' histograms on the horizontal axis of D,F are always highly 
concentrated, corresponding to the fact that, for the beta fecundities
lottery model, $E_i/C_{i \bs i}$ cannot vary much when $E_i$ and 
$C_{i \bs i}$ are in the upper portions of their distributions. 
Likewise, the ``right'' histogram on E is highly concentrated because
$E_i^\nat/C_{i \bs i}^\nat$ cannot vary much when $E_i^\nat$ and 
$C_{i \bs i}^\nat$ are in the upper portions of their distributions.
In contrast, ``left'' histograms on the horizontal axis of D-F
can be highly variable, though only when $E_i$ and $C_{i \bs i}$
(or $E_i^\nat$ and $C_{i \bs i}^\nat$) are not tightly 
correlated in the left portions of their distributions.
The function pictured with a black line on panels D-F is simply the function
$z \rightarrow \ln(1-\delta+10^z)$ that takes $z=\log_{10}(E_1,C_{1 \bs 1})$
to $\ln(1-\delta+E_1/C_{1 \bs 1})$, representing growth rate for the model.
Storage effects are manifest by this function, because it is flat on the left and 
upward sloping on the right, so highly variable horizontal-axis inputs to the function
can translate into higher average vertical-axis values than can concentrated
horizontal-axis inputs of the same mean; e.g., the histograms labeled ``left'' on
E,F. Red dots indicate the mean values of
each histogram. The average vertical-axis values of the two red dots on each panel
equal the GWR of species $i$ for the model (value displayed numerically).

We now relate Fig. \ref{fig:beta_fecundidities_explanatory} back to ATAs. Coexistence is possible for 
right-tail association (panel F, positive GWR) but impossible for left-tail association (panel D,
negative GWR). Coexistence is possible for the right-tail case essentially
because the points in the ``left'' region in C produce highly variable
annual growth rates, which lead to storage effects; the same effect does 
not occur for the left-tail case because 
points in the ``left'' region on panel A are concentrated on line $L_1$ because of the
strong left-tail association of $E_i$ and $C_{i \bs i}$ on that panel. 
Association, or lack thereof, in right tails makes little difference 
to mean growth rates because $E_i/C_{i \bs i}$ cannot vary much. Overall, ATA effects 
are positive for the right-tail case because strong association between 
$E_i$ and $C_{i \bs i}$ is concentrated in the right tails, where there 
is little potential for storage effects anyway; and so disassociation between
$E_i$ and $C_{i \bs i}$ occurs in the left tails, where there is much potential for
storage effects. For the actual quantification of ATA effects,
a comparison is made to the case of symmetric association 
(panels B, E), as a benchmark. What is assessed by ATA
effects is the degree to which patterns of disassociation between 
$E$ and $C$ align with the portions of $E,C$ distribution for which the potential
for storage effects is high. Larger ATA effects indicate better alignment.

Fig. \ref{fig:lognormal_fecundidities_explanatory} is similar to 
Fig. \ref{fig:beta_fecundidities_explanatory}, but is based on the log-normal 
fecundities lottery model instead of the beta fecundities lottery model. 
The figure illustrates that the potential for storage events is equivalent in
the upper and lower tails of the $EC$ distribution (labelled ``right'' and ``left'',
respectively, on panels A-C). That is why ATA effects are fairly modest for the 
log-normal fecundities lottery model for the both the left- and right-tail
associated cases.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLBqij.pdf}
\caption[Analogous to Fig. \ref{MT-fig:fig4} but making a different choice for $q_{ij}$]
{Analogous to Fig. \ref{MT-fig:fig4} of the main text, but using 
$q_{ij}=\eta_i/(\eta_j(1-\delta)+\eta_i \delta)$. See the caption of
that figure for details. Standard errors of plotted quantities 
were never greater than \Sexpr{round(se4qij,5)}, so are not displayed.}\label{fig:fig4_alt}
\end{figure}

@@

\begin{figure}
\includegraphics[width=.66\textwidth]{../results_figs/VKQfig.pdf}
\caption[Maximum growth rate, cellular silicate content, and half saturation constant dependencies on temperature.]
{Species maximum population growth rates, $V_i$ (a), half saturation constants for the dependence 
of growth rates on nutrient concentration, $K_i$ (b), and cellular silicate content, $Q_i$ (c) dependend on 
temperature in ways explored by \cite{Descamps_2005} and \cite{Ellner_2016, Ellner_2019} and 
pictured here. Species 1 is \emph{F. crotonensis}, shown in blue, and species 2 is 
\emph{C. pseudostelligera}, shown in red. Open circles and dashed lines indicate estimated 
values and interpolations from \cite{Descamps_2005} and closed circles and solid lines indicate 
values and functions used by \cite{Ellner_2016, Ellner_2019} and in our analyses. The two sets of
estimates and functions coincide on (c), except for the right-most point for species 2. For that species,
data were not available for $Q$ at $24^\circ$C from the original study of \cite{Descamps_2005}; \cite{Ellner_2016, Ellner_2019}
assumed constancy of $Q_2$ for temperature between $18^\circ$C and $24^\circ$C. Data were also not available at
$24^\circ$C for $K_2$, and \cite{Ellner_2016, Ellner_2019} assumed constancy of $K_2$ for temperatures from 
$12^\circ$C through $24^\circ$C. \label{fig:VKQ}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/PostHocExplanatoryFig_betaFecund.jpg}
\caption[A detailed look at ATA effects for the beta fecundities lottery model]{A 
detailed look at ATA effects for the beta fecundities lottery model.
The figure gives a pictorial representation of the calculation of  
$\E \left[ \ln \left( 1-\delta+\frac{E_i}{C_{i \bs i}} \right) \right]$ (A, C, D, F) and 
$\E \left[ \ln \left( 1-\delta+\frac{E_i^{||}}{C_{i \bs i}^{||}} \right) \right]$ (B, E)
for the model, considering both left- (A, D) and right-tail (C, F) associations
between $E_i$ and $C_{i \bs i}$. ATA effects are the difference of these two values.
The figure also illustrates how ATAs reflect the degree of alignment between where 
in the $EC$ distributions associations are allocated and which parts of that distribution
have high potential for storage effects. See SI section \ref{SIsect:intuition_examples}
for details. We used model parameters $\eta_1 = 1$, $\eta_2 = 1.2$, and $\delta=0.6$
for this example.}\label{fig:beta_fecundidities_explanatory}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/PostHocExplanatoryFig_lognormalFecund.jpg}
\caption[A detailed look at ATA effects for the log-normal fecundities lottery model]{A 
detailed look at ATA effects for the log-normal fecundities lottery model.
Similar to Fig. \ref{fig:beta_fecundidities_explanatory}, but for the 
log-normal fecundities lottery model instead of the beta fecundities
lottery model. See SI section \ref{SIsect:intuition_examples}
for details. We used model parameters $\sigma = 1$, $\mu_1 = 0$, $\mu_2=0.5$, 
and $\delta=0.6$ for this example.}\label{fig:lognormal_fecundidities_explanatory}
\end{figure}
